# Data transformation
```{r, echo=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(sqldf)
school = read.csv("School.csv")
```

## Removing Incomplete Schools
Before removing incomplete schools, the following table shows the number of schools for number of observation occurrences:

| Number of observations |      Number of schools       |
| ---------------------- | ---------------------------- |
|          1 observation |             19 school        |
|          2 observations|             18 schools       |
|          3 observations|             33 schools       |
|          4 observations|             40 schools       |
|          5 observations|            1718 schools      |

Based on the exploratory data analysis result we obtained in the last chapter, we know that there are schools that does not have observation for each of the five academic years, which complicates analysis process if we want to conduct time series analysis. There are 1834 schools included in the data set and 1718 of them have observations for all five years, which constitutes over 93% of all schools. Thus, we choose to directly remove those schools that do not have full observations.
```{r ch3-1}
school <- school %>%
  group_by(School.Name) %>%
  mutate(count = n()) %>%
  filter(count == 5) %>%
  select(-count)
```

## Adding features
### Location of School
Since the first column specifies `DBN(District Borough Number)` of each school, we firstly use it to generate a new `location` variable. According to the DBN assigning system, the Borough is the capitalized letter in the string. `K` represents Brooklyn district, `X`represents the Bronx district, `Q` represents the Queens district, `M` represents the Manhattan district, and `R` represents the Staten Island district. During transformation, we used `str_detect` function in the `stringr` package to check for ocurrences of each Borough letter and assign the corresponding school with a location identifier in terms of the district.

After specifying the location of school, we can first get a capture of school number in different Borough among different academic years in our data source. The number of schools by year and location are shown below:

```{r ch3-2}
# loc <- ggplot(data = school, aes(x = Year)) + 
#   geom_bar(stat="count",fill = "lightblue")+
#   facet_wrap(~location,ncol =5)+
#   ggtitle("Number of schools in different Borough by year") + 
#   theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14), 
#         axis.text.x = element_text(angle = 30, vjust = 0.5))+
#    xlab("Academic Year")+
#   ylab("Number of schools")
# loc
# 
# dis <-  ggplot(data = school, aes(x = Year,fill = location)) + 
#   geom_bar(stat="count") +
#   ggtitle("2013-2018 Time Series School Count") +
#   theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14)) + 
#   xlab("Academic Year")+
#   ylab("Number of schools") 
# dis
```

```{r ch3-3}
school <- school %>% 
  mutate(location = case_when(
    str_detect(DBN, "K") == TRUE ~ "Brooklyn",
    str_detect(DBN,"X") == TRUE ~ "Bronx",
    str_detect(DBN,"Q") == TRUE ~ "Queens",
    str_detect(DBN,"M") == TRUE ~ "Manhattan",
    str_detect(DBN,"R") == TRUE ~ "Staten Island",
  ))

# school %>%
#   ggplot(mapping = aes(x = location)) +
#   geom_bar(stat = "count", fill = "lightblue") +
#   ggtitle("School Count by District") +
#   xlab("Districts") + ylab("Count") +
#   theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))

school %>%
  filter(Year == "2017-18") %>% 
  group_by(location) %>%
  summarise(Count = n()) %>%
  ggplot(mapping = aes(x = location, y = Count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  ggtitle("School Count by District") +
  xlab("Districts") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14)) +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5), 
            size = 3,color = "grey48")
```

### Type of School
As previously mentioned in Chapter2, we can use the number of students enrolled in different grades to judge the type of school. In the dataset, the number of students in different grade are measured from `Grade PK` to `Grade 12`. We use the criteria below to classify the type of School.

| Type of School      | Grades Included              |
| ------------------- | ---------------------------- |
| Pre School Learning | `Grade PK`                   |
| Elementary School   | From `Grade k` to `Grade 5`  |
| Middle School       | From `Grade 6` to `Grade 8`  |
| High School         | From `Grade 9` to `Grade 12` |

```{r ch3-4}
school <- rename(school, Grade.PK =Grade.PK..Half.Day...Full.Day. )
```

```{r ch3-5}
school <- school %>% 
  mutate(is_preschool = case_when(
    Grade.PK > 0 ~ 1,
    T ~ 0
  )) %>% 
  mutate(is_elementary = case_when(
    Grade.K == 0 & Grade.1 == 0 & Grade.2 == 0 &
      Grade.3 == 0 & Grade.4 == 0 & Grade.5 == 0 ~ 0,
    T ~ 1
  )) %>% 
  mutate(is_middle = case_when(
    Grade.6 == 0 & Grade.7 == 0 & Grade.8 == 0 ~ 0,
    T ~ 1
  )) %>% 
  mutate(is_high = case_when(
    Grade.9 == 0 & Grade.10 == 0 & Grade.11 == 0 & Grade.12 == 0 ~ 0,
    T ~ 1
  ))
```

The plot below shows the distribution of different school types at the five identified school districts:

```{r ch3-6}
count_by_type <- sqldf("select location, Year, sum(is_preschool) as Pre_Elementary,
      sum(is_elementary) as Elementary,
      sum(is_middle) as Middle,
      sum(is_high) as High from school group by location ,Year") %>% 
   pivot_longer(-c(location,Year),names_to ="type", values_to = "num")

count_by_type$type <- factor(count_by_type$type ,levels = c("Pre_Elementary", "Elementary", "Middle", "High"))

type_sch <- count_by_type %>%
  ggplot(mapping = aes(x = type, y = num, fill = location)) +
  geom_bar(stat = "identity", position = "fill") +
  ggtitle("School Type Distribution") +
  xlab("School Type") + ylab("Proportion") +
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))+
  scale_fill_brewer(palette = "Pastel2") 
type_sch
```

From the side by side proportion bar plot, we can see that the relative proportion of school types that each district have is quite stable. In other words, we can say that no school district have a preference for a certain school type, and the accessibility for all school levels is equal within each school district, regardless of inter-district difference in the prevalence of school types.

```{r ch3-7}
# type_sch <- ggplot(count_by_type, aes(fill=type, y=num, x=Year)) + 
#     geom_bar(position="dodge", stat="identity")+
#   ggtitle("Number of schools in different type by year") + 
#   theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))+
#    xlab("Academic Year")+
#   ylab("Number of schools") +
#   scale_color_brewer(palette="Dark2")
# type_sch
```


```{r ch3-8}
#write.csv(school, "school_f.csv", row.names = FALSE)
```