# Missing values
```{r, echo=FALSE}
library(knitr)
library(ggplot2)
setwd("~/Desktop")
NYC <- read.csv("NYC.csv")
```
We first check if there are any NA values in the dataset:
```{r}
any(is.na(NYC))
```
However, after closer inspection, we saw that there are entries labeled with "No Data". Thus, we need to treat those entries as NAs.
```{r}
NYC_na <- NYC %>%
  na_if("No Data")
any(is.na(NYC_na))
sum(is.na(NYC_na))
sum(is.na(NYC_na))/prod(dim(NYC_na))*100 #in percentage
```
## Missing Pattern plot
```{r}
missing_plot <- function(data, display = "counts"){
  #missing pattern
  missing_patterns <- data.frame(is.na(data)) %>%
    group_by_all() %>%
    count(name = "count", sort = TRUE) %>%
    ungroup()
  
  key_order <- sort(colSums(data.frame(is.na(data))), decreasing = T)
  key_missing <- rep(key_order, each = nrow(missing_patterns))
  if(display == "counts"){
    #top plot
    up <- key_order %>%
      as.data.frame() %>%
      rownames_to_column("variable") %>%
      ggplot(mapping = aes(x = reorder(variable, desc(.)), y = .))+
      geom_bar(stat = "identity") +
      xlab("") + ylab("count") +
      theme(axis.text.x = element_text(angle = -45, hjust = 0))
    #right plot
    right <- missing_patterns %>%
      rownames_to_column("pattern") %>%
      ggplot(mapping = aes(x = count, y = reorder(pattern, desc(as.numeric(pattern))))) +
      geom_bar(stat = "identity") +
      xlab("row count") + ylab("")
  }
  else if(display == "percent"){
    #top plot
    up <- key_order %>%
      as.data.frame() %>%
      mutate(percentage = ./nrow(data)*100) %>%
      rownames_to_column("variable") %>%
      ggplot(mapping = aes(x = reorder(variable, desc(percentage)), y = percentage))+
      geom_bar(stat = "identity") +
      xlab("") + ylab("% rows missing") +
      ylim(0, 100) +
      theme(axis.text.x = element_text(angle = -45, hjust = 0))
    #right plot
    right <- missing_patterns %>%
      mutate(percentage = count/nrow(data)*100) %>%
      rownames_to_column("pattern") %>%
      ggplot(mapping = aes(x = percentage, y = reorder(pattern, desc(as.numeric(pattern))))) +
      geom_bar(stat = "identity") +
      xlab("%rows") + ylab("") +
      xlim(0, 100)
  }
  
  #identify complete case
  row <- rowSums(missing_patterns[-ncol(missing_patterns)])
  comp_ind <- which(row == 0)
  
  #initialize fill/alpha vector
  fill_alpha <- rep(0.5, nrow(missing_patterns))
  fill_alpha[comp_ind] <- 1
  
  #main plot
  main <- missing_patterns %>% 
    select(-count) %>%
    select(names(key_missing)) %>%
    rownames_to_column("pattern") %>%
    gather(key, value, -pattern) %>%
    mutate(missing = ifelse(value == 0, F, T)) %>%
    cbind(key_missing) %>%
    ggplot(mapping = aes(x = reorder(key, desc(key_missing)), 
                         y = reorder(pattern, desc(as.numeric(pattern))), 
                         fill = missing)) +
    geom_tile(color = "white", show.legend = F, aes(alpha = pattern)) +
    xlab("variable") + ylab("missing pattern") +
    scale_fill_manual(values = c("gray", "purple")) +
    scale_alpha_manual(values = fill_alpha) +
    geom_text(aes(x = names(key_order)[length(key_order)/2+1], 
                  y = as.character(comp_ind), label = "complete cases")) +
    theme(axis.text.x = element_text(angle = -45, hjust = 0))
  
  layout <- "
  BBB##
  AAAC#
  AAAC#
  AAAC#
  "
  main + up + right + plot_layout(design = layout)
}
```
Given the shear size of the dataset, it is hard to visualize missing value pattern of the whole dataset, thus, we can sample from the whole data set and visualize the missing pattern:
```{r}
set.seed(1234567)
# case1
row_ind <- sample(1:nrow(NYC_na), 200)
col_ind <- sample(1:ncol(NYC_na), 10)
new_NYC <- NYC_na[row_ind, col_ind]
missing_plot(new_NYC)

# case2
row_ind <- sample(1:nrow(NYC_na), 200)
col_ind <- sample(1:ncol(NYC_na), 10)
new_NYC <- NYC_na[row_ind, col_ind]
missing_plot(new_NYC)

# case3
row_ind <- sample(1:nrow(NYC_na), 200)
col_ind <- sample(1:ncol(NYC_na), 10)
new_NYC <- NYC_na[row_ind, col_ind]
missing_plot(new_NYC)
```
After trying few cases of missing pattern combinations, it is really rare to find a case that have a pattern. This is because that the data itself has a really large size, which is 8972 rows by 39 columns, while there's only 1730 missing entries, with a missing value rate less than 0.5%. Thus, we can say that the data set is overall complete.
