# Missing values
```{r, echo=FALSE}
library(knitr)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(patchwork)
NYC <- read.csv("School.csv")
```
We first check if there are any NA values in the dataset:
```{r, echo = TRUE}
#Check if there are any NA values in the dataset
any(is.na(NYC))
```
However, after closer inspection, we saw that there are entries labeled with "No Data" in the column `Economic.Need.Index`. Thus, we need to treat those entries as `NA`s. After substitution "No Data" with NA, we calculate the count of missing value and the rate of missing as below:
```{r}
NYC_na <- NYC %>%
  na_if("No Data")
# count of NAs in the dataset
sum(is.na(NYC_na))

# percentage of NA in the column Economic.Need.Index
sprintf("%0.2f%%",100*sum(is.na(NYC_na))/prod(dim(NYC_na))) #in percentage
```
## Missing Pattern plot
```{r}
plot_missing <- function(data, percent = FALSE ){
  missing_patterns <- data.frame(is.na(data)) %>%
    group_by_all() %>%
    count(name = "count", sort = TRUE) %>%
    ungroup() %>%
    remove_rownames() %>%
    rownames_to_column("number") 
  
  missing_patterns_graph<- missing_patterns %>%
    gather(key,value,-c(number,count)) %>%
    group_by(number) %>%
    mutate(complete_case = factor(ifelse(sum(value) == 0, 1, 0.5)))

  if(percent){
    missing_patterns_number <-  missing_patterns_graph %>% 
      group_by(key)%>%
      summarise(missing_number = 100*((sum(value*count))/sum(count)))%>%
      arrange(desc(missing_number)) 
  }
  if(!percent){
    missing_patterns_number <-  missing_patterns_graph %>% 
      group_by(key)%>%
      summarise(missing_number = (sum(value*count))) %>%
      arrange(desc(missing_number))
  }
  
  level_for_order <- (missing_patterns_number$key)

  g1 <- missing_patterns_graph %>%
    ggplot(aes(x = factor(key,level_for_order), 
             y = fct_rev(factor(as.numeric(number))),
             fill = value,alpha = complete_case)) +
    geom_tile(color = "white") + 
    scale_fill_manual(values = c("grey","purple")) +
    scale_alpha_manual(values = c(1,0.5)) +
    annotate("text",x = length(level_for_order)/3,
           y = missing_patterns_graph[missing_patterns_graph$complete_case == 1,]$number,
           label = "Complete Case") +
    theme(legend.position="none",axis.text.x = element_text(angle = -30, hjust = 0)) +
    xlab("variable")+
    ylab("missing pattern")

   
  g2 <- missing_patterns %>% 
    group_by(number)%>%
    ggplot(aes(x = reorder(number,count),y = count)) +
    geom_bar(stat="identity") + 
    ylab("row count")+
    coord_flip()+
    theme(axis.title.y=element_blank()) 
 

  g3 <- ggplot(missing_patterns_number,
               aes(x = reorder(key,-missing_number),y = missing_number)) +
    geom_bar(stat="identity") +
    theme(axis.title.x=element_blank(),axis.title.y = element_text(size = 8),axis.text.x = element_text(angle = -30, hjust = 0)) +
    ggtitle("Missing value patterns")

  
  if(percent){
    g3 <- g3 +  ylab("% rows missing") + ylim(0,100)
  }else{
    g3 <- g3 +  ylab("num rows missing")
  }

  g3 + plot_spacer() + g1 +g2+ plot_layout(widths = c(5, 2), 
                                           heights = unit(c(2, 5), c('cm', 'null')))
}
```
Given the shear size of the dataset, it is hard to visualize missing value pattern of the whole dataset, thus, we can sample from the whole data set and visualize the missing pattern:
```{r}
set.seed(1234567)
# case1
row_ind <- sample(1:nrow(NYC_na), 200)
col_ind <- sample(1:ncol(NYC_na), 10)
new_NYC <- NYC_na[row_ind, col_ind]
plot_missing(new_NYC)
# case2
row_ind <- sample(1:nrow(NYC_na), 200)
col_ind <- sample(1:ncol(NYC_na), 10)
new_NYC <- NYC_na[row_ind, col_ind]
plot_missing(new_NYC)
# case3
row_ind <- sample(1:nrow(NYC_na), 200)
col_ind <- sample(1:ncol(NYC_na), 10)
new_NYC <- NYC_na[row_ind, col_ind]
plot_missing(new_NYC)
```
After trying few cases of missing pattern combinations, it is really rare to find a case that have a pattern. This is because that the data itself has a really large size, which is 8972 rows by 39 columns, while there's only 1730 missing entries, with a missing value rate less than 0.5%. Thus, we can say that the data set is overall complete.
