# Results

```{r, echo=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(sqldf)
library(patchwork)
library(d3r)
library(parcoords)
library(GGally)
library(wesanderson)
library(patchwork)
school = read.csv("school_f.csv")
school <- school %>%
  group_by(School.Name) %>%
  mutate(count = n()) %>%
  filter(count == 5) %>%
  select(-count)
```

## Deep Dive of School Type

### Combination of School Type 
Many schools have multiple levels of educations based on our criteria in Chapter 3. For example, some schools may have both pre-school and elementary school, and combinations of middle school and high school are also common. Therefore, we further constructed another factor variable `type_str` to specify all possible combinations of school types.

The number of schools count by their complete type is distributed as follow: 
```{r chunk5.1}
# type_c <- school %>% 
#   mutate(type_count = is_elementary + is_preschool + is_middle + is_high) %>% 
#   mutate(type_str = case_when(
#     type_count == 1 & is_elementary == 1 ~ "ele",
#     type_count == 1 & is_preschool == 1 ~ "pre",
#     type_count == 1 & is_middle == 1 ~ "mid",
#     type_count == 1 & is_high == 1 ~ "high",
#     type_count == 4 ~ "pre_ele_mid_high",
#     type_count == 2 & is_elementary + is_high == 2 ~ "ele_high",
#     type_count == 2 & is_elementary + is_preschool == 2 ~ "pre_ele",
#     type_count == 2 & is_elementary + is_middle == 2 ~ "ele_mid",
#     type_count == 2 & is_middle + is_high == 2 ~ "mid_high",
#     type_count == 3 & is_elementary + is_preschool + is_middle == 3 ~ "pre_ele_mid",
#     type_count == 3 & is_elementary + is_high + is_middle == 3 ~ "ele_mid_high",
#     T ~ "unknown"
#   )) %>% 
#   filter(Year == "2017-18")

type_c <- school %>% 
  mutate(type_count = is_elementary + is_preschool + is_middle + is_high) %>% 
  mutate(type_str = case_when(
    type_count == 1 & is_elementary == 1 ~ "Elementary School",
    type_count == 1 & is_preschool == 1 ~ "Pre-elementary School",
    type_count == 1 & is_middle == 1 ~ "Middle School",
    type_count == 1 & is_high == 1 ~ "High School",
    type_count == 4 ~ "Pre-elementary, Elementary, Middle, High School",
    type_count == 2 & is_elementary + is_high == 2 ~ "Elementary, High School",
    type_count == 2 & is_elementary + is_preschool == 2 ~ "Pre-elementary, Elementary School",
    type_count == 2 & is_elementary + is_middle == 2 ~ "Elementary, Middle School",
    type_count == 2 & is_middle + is_high == 2 ~ "Middle, High School",
    type_count == 3 & is_elementary + is_preschool + is_middle == 3 ~ "Pre-elementary, Elementary, Middle School",
    type_count == 3 & is_elementary + is_high + is_middle == 3 ~ "Elementary, Middle, High School",
    T ~ "unknown"
  )) %>% 
  filter(Year == "2017-18")
```

```{r chunk5.2}
# type_c$type_str <- factor(type_c$type_str ,levels = c("pre","ele","mid","high","pre_ele","ele_mid","mid_high","ele_high", "pre_ele_mid","ele_mid_high","pre_ele_mid_high"))
type_c$type_str <- factor(type_c$type_str, 
                          levels = c("Pre-elementary School",
                                     "Elementary School", "Middle School",
                                     "High School", 
                                     "Pre-elementary, Elementary, Middle, High School",
                                     "Elementary, High School", 
                                     "Pre-elementary, Elementary School",
                                     "Elementary, Middle School",
                                     "Middle, High School",
                                     "Pre-elementary, Elementary, Middle School",
                                     "Elementary, Middle, High School"
                                     ))
complete_type <- ggplot(type_c, aes(x = fct_rev(type_str)))+
  geom_bar(stat = "count", fill = "lightblue") + 
  coord_flip() +
  xlab("Type of School") +
  ylab("Number of School") +
  ggtitle("School Type Distribution") +
  
  geom_text(stat='count',aes(label=..count..),  hjust =-0.4, color = "grey68") +
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
complete_type
```

* Clearly, the dataset doesn't include schools that only provide pre-school education or there may not be demographic schools in New York City provide only pre-school learning. This phenomenon indicates that all preschools observed and collected in the data set are provided with other levels of educations. Pre-schoolers are the most vulnerable in terms of forming the sense of society, especially equality. Thus, if the pre-schoolers are educated in an environment with other students that are more mentally mature and have more developed social conceptions, then it is possible for those students from higher grades to shape the world-view of the pre-schoolers. In this case, school officials need to provide pre-schoolers with a safe environment to study and form the correct world-views.
* Generally speaking, the combination of school type is based on chronological order of grades. As we can see, the largest quantity of schools have combinations of pre-school learning and elementary school learning part. But surprisingly, there exists a sample of schools with combinations of elementary part and high school part. The unexpected combination of elementary school with high school is alarming to education officials, since high school students are more mature and have relatively developed social concepts, including socioeconomic status differences. On the other hand, students in elementary school barely have any social experiences, let alone the concept of discrimination, thus their perception can be easily compromised when exposed to a discriminated environment. School officials should pay attention to the age differences in student groups, preventing the high school students from bring negative influences to elementary school students. In sum, it is important to provide elementary school students with a healthy environment to grow up.
* The dataset also contains 13 sample of schools which cover from pre-learning to high school grade. We will consider it together with the next part when investigating special schools provided for students with disability.

### Special School for Students with Disability

We use the metric of `proportion of students with disability` to determine whether a school can be considered as a speical school for students with disability.

```{r chunk5.3}
dis_school <- type_c  %>%
  summarise(avg_dis = mean(`X..Students.with.Disabilities.1`),
            quan75_dis = quantile(`X..Students.with.Disabilities.1`)[4])
dis <- ggplot(type_c, aes(x = `X..Students.with.Disabilities.1`)) +
  geom_histogram(fill = "#9fc29f",binwidth = 5) +
  xlab("propotion of student with disability (%)")+
  ylab("Number of schools")+
  ggtitle("Distribution of Propotion of Student with Disability") +
  scale_color_manual(name = "Statistics", values = c(mean = "#80593D")) +
  geom_vline(aes(xintercept = mean(dis_school$avg_dis), color = "mean")) +
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
dis
```

The histogram is right skewed, and with a peak around 100%. Most of demographic schools in New York City have less than 50% students with disability and the mean proportion is about 22%. Based on the distribution, we decide to set $50%$ as a threshold to determine whether the school is a special school for students with disability. To explore more about this special type of schools, we focused on the grade type they provided inspired by the prior part.

```{r chunk5.4}
scatter <- ggplot(type_c, aes(x = type_str, y = `X..Students.with.Disabilities.1`)) +
  geom_point(color = "#2c7bb6",alpha = 0.5)+
  coord_flip() +
  xlab("School Type")+
  ylab("proportion of student with disability(%)")+
  geom_hline(size = 1,aes(yintercept =mean(50), color = "Threshold")) +
  scale_color_manual(name = "Statistics", values = c(Threshold = "#abd9e9"))+
  ggtitle("Student with Disability vs School Type") +
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
scatter
```

* Types of special schools for student with disability are also various. No special elementary school and middle school exists in the dataset, which indicates students with disability in New York City at this stage can involved in normal education together with their peers. Although it is good for all students, regardless of their physical/mental capacities, to have access to the same education, it is still essential to ensure that students who are disabled are treated equally by other students and are protected from possible discrimination.

* The special school surely contributed more sample of schools with full range from pre learning to high school learning. From another perspective, they ensure the equity of achieving all stages of educations for students with disability. 

## Analysis of school enrollment

### Breakdown by location

The distribution of school enrollment by location is shown below. 

```{r ch5.5, warning = F,fig.width=13, fig.height=4}
enroll <- type_c %>% 
  filter(Year == "2017-18")

enroll_dis <- ggplot(enroll, aes(x = `Total.Enrollment`)) + 
  geom_histogram(fill = "lightblue",binwidth = 200) +
  scale_x_continuous(limits = c(0, 6000)) +
  facet_wrap(~location, ncol = 5)+
  xlab("Total number of enrollment")+
  ggtitle("Distribution of School Enrollment by location") +
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
enroll_dis
```

Most of schools in New York City have the number of students varies under quantity of 2000. Outliers of "large" schools which have students more than 2000 are more often located in Brooklyn and Queens, given they both have larger area and populations compared with the other three boroughs. With the notable outliers, with particularly large number of student population, in the aforementioned school districts, there might be a higher risk of school bullying/discrimination. We will further confirm the hypothesis in later section that analyzes race and ethnicity distributions.

### Breakdown by School Type

Since the type of number of grade a school includes can affect its student enrollment, we constructed boxplot to explore the enrollment of different type of schools.

```{r ch5.6}
box1 <- ggplot(enroll, aes(x = fct_reorder(type_str,`Total.Enrollment`), y = `Total.Enrollment`)) +
  geom_boxplot(color = "#117864", fill = "#48c9b0") + 
  coord_flip() +
  xlab("School Type")+
  ggtitle("Boxplot of Total enrollment by School Type") + 
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14)) 
box1
```

* Apart from schools with coverage from pre-learning to high school learning (as stated before, most of them are special school for disability students), we can safely conclude that school with more grade level will have larger student enrollment in average. Therefore, schools provide elementary, middle school and high school level learning are generally have largest quantity of total enrollment since they includes 12 levels of grade.

* Outlier exist in several types, but high school type consists largest amount of outliers, and also tends to have extremely large number of enrollment, expanding from 1000 students to almost 6000 students. This disparity could be due to the difference between education qualities. Because the quality of high school will directly impact the college application of students, a good private high school can provide students with much enriched education environment and allow them to build a strong application background, while a poorer public high school that has lots of students have limited education resources. The large quantity of outliers for High schools is a significant indication of socioeconomic status differences between students.

## School Diversity in NYC

### Race and Ethnicity

When it comes to diversity, race and ethnicity becomes the most popular topic. Here we use percentage bar plot and parallel coordinate plot to see whether there are correlations patterns in race group of public schools in New York City.

```{r mosaic}
 
school <- school %>% rename(
  total_enroll = `Total.Enrollment`,asian = `X..Asian`,
                            black = `X..Black`,
                            hispanic = `X..Hispanic`,
                            white = `X..White`,
                            others = `X..Multiple.Race.Categories.Not.Represented`)
```

```{r chunk race}
race_time_series <-school %>% 
    group_by(location) %>%
    summarise(Asian = sum(asian)/sum(total_enroll),
              Black = sum(black)/sum(total_enroll),
              Hispanic = sum(hispanic)/sum(total_enroll),
              White = sum(white)/sum(total_enroll),
              Others = sum(others)/sum(total_enroll))%>%  pivot_longer(-location,names_to = "Race",
  values_to = "proportion")
```

```{r patchwork}
race_all <- sqldf("select sum(asian) asian, sum(black) black, sum(hispanic) hispanic, sum(white) white,sum(others) others,sum(total_enroll) total from school") %>% pivot_longer(-total, names_to = "Race",
  values_to = "proportion") %>% 
  mutate(percent = proportion / total) %>% 
  mutate(total = "NYC") %>% 
  select(-proportion)

all_race <- ggplot(race_all, aes(x = total, y = percent, fill=Race))+
geom_bar(stat="identity",position="fill")+
  scale_fill_brewer(palette = "Pastel1") +
  xlab(" ")+
  ylab("Proportion(%)")+
geom_text(aes(label = round(percent*100,1)), position = position_stack(vjust = 0.5), size = 3,color = "grey48") +
  theme(legend.position="none",plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
  
race_plot <- ggplot(race_time_series, aes(x = location, y = proportion, fill=Race))+
geom_bar(stat="identity",position="fill")+
geom_text(aes(label = round(proportion*100,1)), position = position_stack(vjust = 0.5), size = 3,color = "grey48") +
scale_fill_brewer(palette = "Pastel1") +
  ylab("")

all_race + race_plot + plot_layout(widths = c(2, 5))+
ggtitle("Proportion of Race in School by Location") +
theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
```

From the percentage bar plot above, we can get a complete view of Race diversity and also its relationship to the location:

* In total, the percentage of Black and Hispanic students are overweighted than White students. According to [NYC Race and Ethnicity data](https://datausa.io/profile/geo/new-york-ny) in 2021, more than 31.9 residents in NYC self-identified as white and non-Hispanic, while the majority of students in public schools are Hispanic and black. 

* Breakdown of locations give us more insights about race and ethnicity issue. The composition of different ethnic groups clearly depends on the school district where the observed school locates in. For example, the proportion of white students enrolled in Staten Island takes almost half of the enrollment, while in other Borough especially Bronx, white students only constitute 4.2% of the total enrollment. This location-dependent ethnicity distribution indicates potential factors, such as social economical discrepancies, that drives families to send their children to a specific school district for school. This disproportionate distribution of ethnicity groups in different districts could serve as a warning to officials about possible discrimination to the minorities in the school. Thus, local agencies and school representatives need to take precaution about potential racial discrimination.

Below we create a random sample of school to create a parallel coordinate plot. You can select, drag and reorder the columns to get a glimpse of race and ethnicity diversity in New York City public school.

**Hint:** Try to select sample with higher percentage of Black students, what's the pattern considering other race group and location?

```{r parcood}
school <- read.csv("school_f.csv")
sample <- school %>% 
  select(`School.Name`, `X..Asian.1`,`X..Black.1`,`X..White.1`,`X..Hispanic.1`,`X..Multiple.Race.Categories.Not.Represented.1`,location) %>% 
  rename(asian = `X..Asian.1`,
                            black = `X..Black.1`,
                            hispanic = `X..Hispanic.1`,
                            white = `X..White.1`,
                            others = `X..Multiple.Race.Categories.Not.Represented.1`) %>% 
  slice_sample(n = 20) 
parcoords(sample,rownames = F 
    , brushMode = "1D-axes"
    , reorderable = T
    , queue = T
    , color = list(colorBy = "location", colorScale = "scaleOrdinal"
      ,colorScheme = "schemeCategory10"),withD3=TRUE
    )   
```

### School Type and Ethnicity Composition
Given the location-dependent ethnicity distribution we discovered in the last section, we continued explore other indicators that may segregate different ethnic groups. The graph below shows the distribution of different ethnicity groups in different school types, where the school types were defined in a previous section.

```{r ch5.sch_eth}
school <- read.csv("School.csv")
school <- school %>%
  #remove incomplete schools (not surveyed in all 5 years)
  group_by(School.Name) %>%
  mutate(count = n()) %>%
  filter(count == 5) %>%
  select(-count) %>%
  ungroup() %>%
  #assign location (according to DBN code)
  mutate(location = case_when(
    str_detect(DBN, "K") == TRUE ~ "Brooklyn",
    str_detect(DBN,"X") == TRUE ~ "Bronx",
    str_detect(DBN,"Q") == TRUE ~ "Queens",
    str_detect(DBN,"M") == TRUE ~ "Manhattan",
    str_detect(DBN,"R") == TRUE ~ "Staten Island",
  )) %>%
  #reformatting column names & removing unwanted columns 
  rename(Grade.PK =Grade.PK..Half.Day...Full.Day. ) %>%
  mutate(is_preschool = case_when(
    Grade.PK > 0 ~ 1,
    T ~ 0
  )) %>% 
  mutate(is_elementary = case_when(
    Grade.K == 0 & Grade.1 == 0 & Grade.2 == 0 &
      Grade.3 == 0 & Grade.4 == 0 & Grade.5 == 0 ~ 0,
    T ~ 1
  )) %>% 
  mutate(is_middle = case_when(
    Grade.6 == 0 & Grade.7 == 0 & Grade.8 == 0 ~ 0,
    T ~ 1
  )) %>% 
  mutate(is_high = case_when(
    Grade.9 == 0 & Grade.10 == 0 & Grade.11 == 0 & Grade.12 == 0 ~ 0,
    T ~ 1
  )) %>%
  #school type assignment
  mutate(type_count = is_elementary + is_preschool + is_middle + is_high) %>% 
  mutate(type_str = case_when(
    type_count == 1 & is_elementary == 1 ~ "Elementary School",
    type_count == 1 & is_preschool == 1 ~ "Pre-elementary School",
    type_count == 1 & is_middle == 1 ~ "Middle School",
    type_count == 1 & is_high == 1 ~ "High School",
    type_count == 4 ~ "Pre-elementary, Elementary, Middle, High School",
    type_count == 2 & is_elementary + is_high == 2 ~ "Elementary, High School",
    type_count == 2 & is_elementary + is_preschool == 2 ~ "Pre-elementary, Elementary School",
    type_count == 2 & is_elementary + is_middle == 2 ~ "Elementary, Middle School",
    type_count == 2 & is_middle + is_high == 2 ~ "Middle, High School",
    type_count == 3 & is_elementary + is_preschool + is_middle == 3 ~ "Pre-elementary, Elementary, Middle School",
    type_count == 3 & is_elementary + is_high + is_middle == 3 ~ "Elementary, Middle, High School",
    T ~ "unknown"
  )) %>%
  select(-c(DBN, X..Female.1, X..Male.1, X..Asian.1, X..Black.1, X..Hispanic.1,
            X..Multiple.Race.Categories.Not.Represented.1, X..White.1,
            X..Students.with.Disabilities.1, X..English.Language.Learners.1,
            X..Poverty.1, is_preschool, is_elementary, is_middle, is_high, type_count)) %>%
  rename(Female = X..Female, Male = X..Male, Asian = X..Asian, Black = X..Black,
         Hispanic = X..Hispanic, Others = X..Multiple.Race.Categories.Not.Represented,
         White = X..White, Disabled.Students = X..Students.with.Disabilities,
         English.Learners = X..English.Language.Learners, Poverty = X..Poverty,
         School.type = type_str)

school %>%
  group_by(School.type) %>%
  summarise(Asian = sum(Asian),
            Black = sum(Black),
            Hispanic = sum(Hispanic),
            White = sum(White),
            Others = sum(Others)) %>%
  pivot_longer(!School.type, names_to = "Ethnicity", values_to = "Count") %>%
  ggplot(mapping = aes(y = School.type, x = Count)) +
  geom_bar(stat = "identity", position = "fill", mapping = aes(fill = Ethnicity), alpha = 0.8) +
  ggtitle("Ethnicity Distribution by School Type") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14)) +
  xlab("Proportion") + ylab("School Type") +
  scale_fill_brewer(palette = "Pastel1")
```

According to the stacked percentage barplot of ethnic group distribution in different school types, we can see high fluctuation of Asian and Black student proportions. Specifically, Black students have higher proportions in schools that are for Elementary, Middle, High Schools and Elementary, High schools, while lower proportions in schools that are identified as only Elementary schools. This difference in ethnicity distribution could indicate that discrepancy among students' families' social economical-status. School that are identified to contain multiple school types are usually public and can be subjected to higher chances of bullying, given the greater distribution of age groups.

### Poverty Level

Poverty is another factor worth considering when talking about school diversity. We first tried to capture a top-down view of `Poverty Level` using histogram. If the students are random selected to be enrolled in demographic schools, the distribution of their poverty level will be nearly normal distributed.

```{r ch5.7}
pover_dis <- ggplot(type_c, aes(x = `X..Poverty.1`,y =..density..)) + 
  geom_histogram(binwidth = 5,fill = "#9FC29F",colour = "#80593D") +
  geom_density(color = "#80593D",size = 0.8) + 
  
  xlab("proportion of students in poverty") +
  facet_wrap(~location, ncol = 3)+
  ylab("")
pover_dis
```

Clearly, the distribution of proportion of poverty students are not normal distributed. By breakdown of location, most of distributions are right skewed except samples from Staten Island, and indicated that a majority of students in demographic schools are experiencing poverty.

### English Language Learner

> English-language learners, or ELLs, are students who are unable to communicate fluently or learn effectively in English, who often come from non-English-speaking homes and backgrounds, and who typically require specialized or modified instruction in both the English language and in their academic courses.

The number and proportion of English learner in public school reflects the diversity in Language and also can be a strong indication of race group and immigration living preference. 

```{r ell}
results <- type_c %>% group_by(location) %>%
            do(model =lm(`X..Asian.1`+`X..Hispanic.1`+ `X..Multiple.Race.Categories.Not.Represented.1` ~ `X..English.Language.Learners.1`, data = .)) %>%
  mutate(coef=coef(model)["X..English.Language.Learners.1"]) %>% 
  select(location,coef) %>% 
  arrange(coef)

ell <- ggplot(type_c, aes(x = `X..Asian.1`+`X..Hispanic.1`+ `X..Multiple.Race.Categories.Not.Represented.1`, y = `X..English.Language.Learners.1`)) +
  xlab("Proportion of minority race student (Asian, Hispanic and Others) (%)") +
  ylab("Proportion of English Language Learner")+
  geom_point(col = "#fdae61", alpha = 0.8)+
  geom_smooth(color = "#b2abd2", method = "lm", se = FALSE) +
  facet_wrap(~factor(location, levels=results$location),ncol = 3)
ell
```

Using the proportion of minority race group vs proportion of English Language Learner, we construct scatter plot with facet of different locations. By definition and clearly in plots, these two factors are positive correlated. Moreover, the sample size and the coefficient of correlation in Staten Island are both the smallest, which may indicate a minority of immigrants in this borough. The slope of the correlations is less than 1, which means that only part of the students are non-native to the English-speaking environment, others are immigrants who grew up in the US. Among the five school districts, Bronx has the largest slope for the regression line, meaning that most of the minority ethnicity students are newly immigrated to the region. The existance of English-learners who are newly immigrated to the region needs special attention from teachers and school officials. Given their limited communication ability, they are more vulnerable to racial discrimination. At the same time, because they are new to the environment, they are helpless when they receive unequal treatments from others.

## Correlation Hypothesis and Analysis

As projected in Chapter 2, we proposed several relationships which can be further investigated through our dataset. In this section, we are going to look at the relationship between factors which affects school diversity given from different perspectives.

```{r}
school <- read.csv("School.csv")
school <- school %>%
  #remove incomplete schools (not surveyed in all 5 years)
  group_by(School.Name) %>%
  mutate(count = n()) %>%
  filter(count == 5) %>%
  select(-count) %>%
  ungroup() %>%
  #assign location (according to DBN code)
  mutate(location = case_when(
    str_detect(DBN, "K") == TRUE ~ "Brooklyn",
    str_detect(DBN,"X") == TRUE ~ "Bronx",
    str_detect(DBN,"Q") == TRUE ~ "Queens",
    str_detect(DBN,"M") == TRUE ~ "Manhattan",
    str_detect(DBN,"R") == TRUE ~ "Staten Island",
  )) %>%
  #reformatting column names & removing unwanted columns 
  rename(Grade.PK =Grade.PK..Half.Day...Full.Day. ) %>%
  mutate(is_preschool = case_when(
    Grade.PK > 0 ~ 1,
    T ~ 0
  )) %>% 
  mutate(is_elementary = case_when(
    Grade.K == 0 & Grade.1 == 0 & Grade.2 == 0 &
      Grade.3 == 0 & Grade.4 == 0 & Grade.5 == 0 ~ 0,
    T ~ 1
  )) %>% 
  mutate(is_middle = case_when(
    Grade.6 == 0 & Grade.7 == 0 & Grade.8 == 0 ~ 0,
    T ~ 1
  )) %>% 
  mutate(is_high = case_when(
    Grade.9 == 0 & Grade.10 == 0 & Grade.11 == 0 & Grade.12 == 0 ~ 0,
    T ~ 1
  )) %>%
  #school type assignment
  mutate(type_count = is_elementary + is_preschool + is_middle + is_high) %>% 
  mutate(type_str = case_when(
    type_count == 1 & is_elementary == 1 ~ "Elementary School",
    type_count == 1 & is_preschool == 1 ~ "Pre-elementary School",
    type_count == 1 & is_middle == 1 ~ "Middle School",
    type_count == 1 & is_high == 1 ~ "High School",
    type_count == 4 ~ "Pre-elementary, Elementary, Middle, High School",
    type_count == 2 & is_elementary + is_high == 2 ~ "Elementary, High School",
    type_count == 2 & is_elementary + is_preschool == 2 ~ "Pre-elementary, Elementary School",
    type_count == 2 & is_elementary + is_middle == 2 ~ "Elementary, Middle School",
    type_count == 2 & is_middle + is_high == 2 ~ "Middle, High School",
    type_count == 3 & is_elementary + is_preschool + is_middle == 3 ~ "Pre-elementary, Elementary, Middle School",
    type_count == 3 & is_elementary + is_high + is_middle == 3 ~ "Elementary, Middle, High School",
    T ~ "unknown"
  )) %>%
  select(-c(DBN, X..Female.1, X..Male.1, X..Asian.1, X..Black.1, X..Hispanic.1,
            X..Multiple.Race.Categories.Not.Represented.1, X..White.1,
            X..Students.with.Disabilities.1, X..English.Language.Learners.1,
            X..Poverty.1, is_preschool, is_elementary, is_middle, is_high, type_count)) %>%
  rename(Female = X..Female, Male = X..Male, Asian = X..Asian, Black = X..Black,
         Hispanic = X..Hispanic, Others = X..Multiple.Race.Categories.Not.Represented,
         White = X..White, Disabled.Students = X..Students.with.Disabilities,
         English.Learners = X..English.Language.Learners, Poverty = X..Poverty,
         School.type = type_str)
```

### Poverty Rate and School Population

Firstly, we care about the single influencer of Poverty level in public schools. Considering the size of school enrollment and poverty rate, we construct scatter plot to see whether there exists patterns, like larger school with high poverty level or vice versa. 

```{r ch5.8}
Poverty.Enrollment <- school %>%
  #remove values that are outliers
  filter(Total.Enrollment <= 1500, Poverty <= 1000) %>%
  group_by(School.Name) %>%
  summarise(Asian = round(mean(Asian), 0),
            Hispanic = round(mean(Hispanic), 0),
            Black = round(mean(Black), 0),
            Others = round(mean(Others), 0),
            White = round(mean(White), 0),
            `Total Enrollment` = round(mean(Total.Enrollment), 0),
            Poverty = mean(Poverty)/ round(mean(Total.Enrollment), 0)) %>%
  ungroup() %>%
  select(-School.Name) %>%
  pivot_longer(!c(`Total Enrollment`, Poverty), names_to = "Ethnicity", values_to = "Count") %>%
  ggplot(mapping = aes(x = `Total Enrollment`, y = Poverty)) +
  geom_point(alpha = 0.5, size = 1,fill = "grey48") +
  geom_density_2d_filled(contour_var = "ndensity",alpha = 0.3, show.legend = FALSE) +
  geom_smooth(method = "lm") +
  ggtitle("Total Enrollment and Poverty Count") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
Poverty.Enrollment

# layout <- "
# AABB
# CCCC
# CCCC
# "
# Poverty + Enrollment + Poverty.Enrollment +
#   plot_layout(design = layout) +
#   plot_annotation(title = "Interaction Between Poverty Count and Total Enrollment",
#                   subtitle = "Contour Map Overlay by Observation Count") +
#   theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

* As the scatter plot shown, while the variance of sample from public schools are large, it is still possible to see the negative relationship between school size and poverty rate. According to the density heat map, the majority of the observations are clustered around Total Enrollment of 250-750, with poverty rate higher than 0.75.

* Surprisingly, from the scatterplot trend, it seems that as the number of students increase in a school, the poverty proportion decreases. This could be due to the buffering effect of a larger school population, or it could be due to the existence of extreme values, such as the one school that has approximately 300 students but nearly 0% poverty. Outliers exist in every dimensions, but there appears to be a straight line border with a negative slope above the collection of points in the plot when total enrollment is larger than 1000. 

* The existence of extreme values of schools that have nearly 0% poverty rate serves as a strong evidence for socioeconomical discrepancies between schools. There is few schools that provide education to only kids from rich families, while the majority of students go to schools that have a poorer background.

```{r ch5.15}
#
school %>%
  group_by(School.Name) %>%
  summarise(Asian = round(mean(Asian), 0),
            Hispanic = round(mean(Hispanic), 0),
            Black = round(mean(Black), 0),
            Others = round(mean(Others), 0),
            White = round(mean(White), 0),
            `Total Enrollment` = round(mean(Total.Enrollment), 0)) %>%
  ungroup() %>%
  select(-School.Name) %>%
  pivot_longer(!`Total Enrollment`, names_to = "Ethnicity", values_to = "Count") %>%
  ggplot(mapping = aes(x = `Total Enrollment`, y = Count)) +
  geom_point(mapping = aes(color = Ethnicity), size = 0.5) +
  geom_smooth(method = "lm") + ylab("Ethnicity Group Count") +
  facet_wrap(~Ethnicity, ncol = 3) +
  scale_fill_brewer(palette = "Pastel1") +
  ggtitle("Total Enrollment and Ethnicity Relationship") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

### Three-way Interaction: Race Composition, Total Enrollment, Poverty

As we hypothesized, ethnic composition may have strong relations with the proportion of students in poverty since the minority ethnic groups are more likely to be immigrants and has lower economics income. 

```{r 5.9}
Asian <- school %>%
  #remove values that are outliers
  filter(Poverty <= 1000) %>%
  group_by(School.Name) %>%
  summarise(Asian = round(mean(Asian), 0),
            Hispanic = round(mean(Hispanic), 0),
            Black = round(mean(Black), 0),
            Others = round(mean(Others), 0),
            White = round(mean(White), 0),
            `Total Enrollment` = round(mean(Total.Enrollment), 0),
            Poverty = mean(Poverty)) %>%
  ungroup() %>%
  select(-School.Name) %>%
  pivot_longer(!c(`Total Enrollment`, Poverty), names_to = "Ethnicity", values_to = "Count") %>%
  filter(Ethnicity == "Asian", Count <= 750) %>%
  ggplot(mapping = aes(x = Poverty, y = Count)) +
  geom_point(aes(color = `Total Enrollment`), alpha = 0.8, size = 1.5) +
  scale_colour_gradient(low = "#93D1FF", high = "#01247E") +
  geom_smooth(method = "lm") + ylab("Asian Count") 
  #geom_density_2d_filled(alpha = 0.5, show.legend = FALSE)
```

```{r ch5.10}
Black <- school %>%
  #remove values that are outliers
  filter(Poverty <= 1000) %>%
  group_by(School.Name) %>%
  summarise(Asian = round(mean(Asian), 0),
            Hispanic = round(mean(Hispanic), 0),
            Black = round(mean(Black), 0),
            Others = round(mean(Others), 0),
            White = round(mean(White), 0),
            `Total Enrollment` = round(mean(Total.Enrollment), 0),
            Poverty = mean(Poverty)) %>%
  ungroup() %>%
  select(-School.Name) %>%
  pivot_longer(!c(`Total Enrollment`, Poverty), names_to = "Ethnicity", values_to = "Count") %>%
  filter(Ethnicity == "Black", Count <= 750) %>%
  ggplot(mapping = aes(x = Poverty, y = Count)) +
  geom_point(aes(color = `Total Enrollment`), alpha = 0.8, size = 1.5) +
  scale_colour_gradient(low = "#93D1FF", high = "#01247E") +
  geom_smooth(method = "lm") + ylab("Black Count")
  #geom_density_2d_filled(alpha = 0.5, show.legend = FALSE)
```

```{r ch5.11}
Hispanic <- school %>%
  #remove values that are outliers
  filter(Poverty <= 1000) %>%
  group_by(School.Name) %>%
  summarise(Asian = round(mean(Asian), 0),
            Hispanic = round(mean(Hispanic), 0),
            Black = round(mean(Black), 0),
            Others = round(mean(Others), 0),
            White = round(mean(White), 0),
            `Total Enrollment` = round(mean(Total.Enrollment), 0),
            Poverty = mean(Poverty)) %>%
  ungroup() %>%
  select(-School.Name) %>%
  pivot_longer(!c(`Total Enrollment`, Poverty), names_to = "Ethnicity", values_to = "Count") %>%
  filter(Ethnicity == "Hispanic", Count <= 750) %>%
  ggplot(mapping = aes(x = Poverty, y = Count)) +
  geom_point(aes(color = `Total Enrollment`), alpha = 0.8, size = 1.5) +
  scale_colour_gradient(low = "#93D1FF", high = "#01247E") +
  geom_smooth(method = "lm") + ylab("Hispanic Count") 
  #geom_density_2d_filled(alpha = 0.5, show.legend = FALSE)
```

```{r ch5.12}
White <- school %>%
  #remove values that are outliers
  filter(Poverty <= 1000) %>%
  group_by(School.Name) %>%
  summarise(Asian = round(mean(Asian), 0),
            Hispanic = round(mean(Hispanic), 0),
            Black = round(mean(Black), 0),
            Others = round(mean(Others), 0),
            White = round(mean(White), 0),
            `Total Enrollment` = round(mean(Total.Enrollment), 0),
            Poverty = mean(Poverty)) %>%
  ungroup() %>%
  select(-School.Name) %>%
  pivot_longer(!c(`Total Enrollment`, Poverty), names_to = "Ethnicity", values_to = "Count") %>%
  filter(Ethnicity == "White", Count <= 750) %>%
  ggplot(mapping = aes(x = Poverty, y = Count)) +
  geom_point(aes(color = `Total Enrollment`), alpha = 0.8, size = 1.5) +
  scale_colour_gradient(low = "#93D1FF", high = "#01247E") +
  geom_smooth(method = "lm") + ylab("White Count")
  #geom_density_2d_filled(alpha = 0.5, show.legend = FALSE)

(Hispanic + Asian )/ (Black + White) +
  plot_annotation(
  title = 'Interaction Between Ethnicity Group and School Poverty',
  subtitle = "Overlay by School Total Enrollment"
)
```

The result from scatter plots of ethic group breakdown supports our hypothesis. Apart from white students proportion, the sample of poverty and ethic group proportion are all positive correlated. Moreover, the slope steepness of positive correlation between poverty and proportion of Hispanic students is largest among four ethic groups, indicates that schools with higher proportion of Hispanic students are more likely to have higher poverty level. 

### Gender Composition and Poverty

Recall our assumption in chapter 2, we try to explore the relationship between school poverty level and gender composition. Based on social conventions, we want to know whether gender bias in education can also be reflected by observations in NYC public schools.

```{r}
gend<-school %>% 
  group_by(School.Name) %>%
  summarise(male_per = round(mean(Male), 0)/  round(mean(`Total.Enrollment`), 0),
            poverty_per = round(mean(Poverty),0) / round(mean(`Total.Enrollment`), 0)) %>% 
  ungroup() %>% 
  select(-School.Name) %>%
  ggplot(mapping = aes(x = poverty_per, y = male_per)) +
  geom_point(alpha = 0.8, size = 1.5) +
  geom_density_2d_filled(contour_var = "ndensity",alpha = 0.3, show.legend = FALSE) +
  xlab("Proportion of Poverty Students in School")+
  ylab("Proportion od Male Students in School")+
  ggtitle("Proportion of Gender vs Poverty Level") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
gend
```

As we shown in poverty level before, most public schools in NYC have large quantity of students in poverty. However, though the poverty level is high, the average proportion of male students in NYC public schools are around 50%, which is a good sign indicating that there's no bias when considering gender diversity and poverty level. 

