# Results

```{r, echo=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(sqldf)
library(patchwork)
library(d3r)
library(parcoords)
library(GGally)
library(wesanderson)
library(patchwork)
school = read.csv("school_f.csv")
school <- school %>%
  group_by(School.Name) %>%
  mutate(count = n()) %>%
  filter(count == 5) %>%
  select(-count)
```

## Deep Dive of School Type

### Combination of School Type 
Many schools have multiple types of grade based on our criteria in Chapter 3. For example, some schools may have both pre-school grade and elementary school grade, and combinations of middle school and high school are also common. Therefore, we construct another factor variable `type_str` to specify all types containing by a school.

The number of schools count by their complete type is distributed as follow: 
```{r chunk5.1}
type_c <- school %>% 
  mutate(type_count = is_elementary + is_preschool + is_middle + is_high) %>% 
  mutate(type_str = case_when(
    type_count == 1 & is_elementary == 1 ~ "ele",
    type_count == 1 & is_preschool == 1 ~ "pre",
    type_count == 1 & is_middle == 1 ~ "mid",
    type_count == 1 & is_high == 1 ~ "high",
    type_count == 4 ~ "pre_ele_mid_high",
    type_count == 2 & is_elementary + is_high == 2 ~ "ele_high",
    type_count == 2 & is_elementary + is_preschool == 2 ~ "pre_ele",
    type_count == 2 & is_elementary + is_middle == 2 ~ "ele_mid",
    type_count == 2 & is_middle + is_high == 2 ~ "mid_high",
    type_count == 3 & is_elementary + is_preschool + is_middle == 3 ~ "pre_ele_mid",
    type_count == 3 & is_elementary + is_high + is_middle == 3 ~ "ele_mid_high",
    T ~ "unknown"
  )) %>% 
  filter(Year == "2017-18")
```

```{r chunk5.2}
type_c$type_str <- factor(type_c$type_str ,levels = c("pre","ele","mid","high","pre_ele","ele_mid","mid_high","ele_high", "pre_ele_mid","ele_mid_high","pre_ele_mid_high"))
complete_type <- ggplot(type_c, aes(x = fct_rev(type_str)))+
  geom_bar(stat = "count", fill = "lightblue") + 
  coord_flip() +
  xlab("Type of School") +
  ylab("Number of School") +
  ggtitle("School Type Distribution") +
  
  geom_text(stat='count',aes(label=..count..),  hjust =-0.4, color = "grey68") +
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
complete_type
```

* Clearly, the dataset doesn't include schools that only provide pre-school education or there may not be demographic schools in New York City provide only pre-school learning. 
* Generally speaking, the combination of school type is based on chronological order of grades. As we can see, the largest quantity of schools have combinations of pre-school learning and elementary school learning part. But surprisingly, there exists a sample of schools with combinations of elementary part and high school part.
* The dataset also contains 13 sample of schools which cover from pre-learning to high school grade. We will consider it together with the next part when investigating special schools provided for students with disability.

### Special School for Students with Disability

We use the metric of `proportion of students with disability` to determine whether a school can be considered as a speical school for students with disability.

```{r chunk5.3}
dis_school <- type_c  %>%
  summarise(avg_dis = mean(`X..Students.with.Disabilities.1`),
            quan75_dis = quantile(`X..Students.with.Disabilities.1`)[4])
dis <- ggplot(type_c, aes(x = `X..Students.with.Disabilities.1`)) +
  geom_histogram(fill = "#9fc29f",binwidth = 5) +
  xlab("propotion of student with disability (%)")+
  ylab("Number of schools")+
  ggtitle("Distribution of Propotion of Student with Disability") +
  scale_color_manual(name = "Statistics", values = c(mean = "#80593D")) +
  geom_vline(aes(xintercept = mean(dis_school$avg_dis), color = "mean")) +
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
dis
```

The histogram is right skewed, and with a peak around 100%. Most of demographic schools in New York City have less than 50% students with disability and the mean proportion is about 22%. Based on the distribution, we decide to set $50%$ as a threshold to determine whether the school is a special school for students with disability. To explore more about this special type of schools, we focused on the grade type they provided inspired by the prior part.

```{r chunk5.4}
scatter <- ggplot(type_c, aes(x = type_str, y = `X..Students.with.Disabilities.1`)) +
  geom_point(color = "#2c7bb6",alpha = 0.5)+
  coord_flip() +
  xlab("School Type")+
  ylab("proportion of student with disability(%)")+
  geom_hline(size = 1,aes(yintercept =mean(50), color = "Threshold")) +
  scale_color_manual(name = "Statistics", values = c(Threshold = "#abd9e9"))+
  ggtitle("Student with Disability vs School Type") +
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
scatter
```

**Remarks:**

* Types of special schools for student with disability are also various. No special elementary school and middle school exists in the dataset, which indicates students with disability in New York City at this stage can involved in normal education together with their peers. 

* The special school surely contributed more sample of schools with full range from pre learning to high school learning. From another perspective, they ensure the equity of achieving all stages of educations for students with disability. 

## Analysis of school enrollment

**By Location:** The distribution of school enrollment by location is shown below. 

```{r ch5.5, warning = F}
enroll <- type_c %>% 
  filter(Year == "2017-18")

enroll_dis <- ggplot(enroll, aes(x = `Total.Enrollment`)) + 
  geom_histogram(fill = "lightblue",binwidth = 200) +
  scale_x_continuous(limits = c(0, 6000)) +
  facet_wrap(~location, ncol = 3)
enroll_dis
```

* Most of schools in New York City have the number of students varies under quantity of 2000. Outliers of "large" schools which have students more than 2000 are more often located in Brooklyn and Queens, given they both have larger area and populations compared with the other three boroughs. 

**By School Type:** Since the type of number of grade a school includes can affect its student enrollment, we constructed boxplot to explore the enrollment of different type of schools.

```{r ch5.6}
box1 <- ggplot(enroll, aes(x = fct_reorder(type_str,`Total.Enrollment`), y = `Total.Enrollment`)) +
  geom_boxplot(color = "#117864", fill = "#48c9b0") + 
  coord_flip() +
  xlab("School Type")+
  ggtitle("Boxplot of Total enrollment by School Type") + 
  theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14)) 
box1
```

* Apart from schools with coverage from pre-learning to high school learning(as stated before, most of them are special school for disability students), we can safely conclude that school with more grade level will have larger student enrollment in average. Therefore, schools provide elementary, middle school and high school level learning are generally have largest quantity of total enrollment since they includes 12 levels of grade.

* Outlier exist in several types, but high school type consists largest amount of outliers, and also tends to have extremely large number of enrollment, expanding from 1000 students to almost 6000 students.

## School Diversity in NYC

### Race and Ethnicity

When it comes to diversity, race and ethnicity becomes the most popular topic. Here we use parallel coordinate plot to see whether there are correlations patterns in race group of public schools in New York City.

```{r mosaic}
 
school <- school %>% rename(
  total_enroll = `Total.Enrollment`,asian = `X..Asian`,
                            black = `X..Black`,
                            hispanic = `X..Hispanic`,
                            white = `X..White`,
                            others = `X..Multiple.Race.Categories.Not.Represented`)
```

```{r chunk race}
race_time_series <-school %>% 
    group_by(location) %>%
    summarise(Asian = sum(asian)/sum(total_enroll),
              Black = sum(black)/sum(total_enroll),
              Hispanic = sum(hispanic)/sum(total_enroll),
              White = sum(white)/sum(total_enroll),
              Others = sum(others)/sum(total_enroll))%>%  pivot_longer(-location,names_to = "Race",
  values_to = "proportion")
```

```{r}
race_all <- sqldf("select sum(asian) asian, sum(black) black, sum(hispanic) hispanic, sum(white) white,sum(others) others,sum(total_enroll) total from school") %>% pivot_longer(-total, names_to = "Race",
  values_to = "proportion") %>% 
  mutate(percent = proportion / total) %>% 
  mutate(total = "NYC") %>% 
  select(-proportion)

all_race <- ggplot(race_all, aes(x = total, y = percent, fill=Race))+
geom_bar(stat="identity",position="fill")+
  scale_fill_brewer(palette = "Pastel1") +
  xlab(" ")+
  ylab("Proportion(%)")+
geom_text(aes(label = round(percent*100,1)), position = position_stack(vjust = 0.5), size = 3,color = "grey48") +
  theme(legend.position="none",plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
  
race_plot <- ggplot(race_time_series, aes(x = location, y = proportion, fill=Race))+
geom_bar(stat="identity",position="fill")+
geom_text(aes(label = round(proportion*100,1)), position = position_stack(vjust = 0.5), size = 3,color = "grey48") +
scale_fill_brewer(palette = "Pastel1") +
  ylab("")

all_race + race_plot + plot_layout(widths = c(2, 5))+
ggtitle("Proportion of Race in School by Location") +
theme(plot.title = element_text(hjust = 0.5, face  = "bold", size = 14))
```

From the percentage bar plot above, we can get a complete view of Race diversity and also its relationship to the location. 

* In total, the percentage of Black and Hispanic students are overweighted than White students. According to [NYC Race and Ethnicity data](https://datausa.io/profile/geo/new-york-ny) in 2021, more than 31.9 residents in NYC self-identified as white and non-Hispanic, while the majority of students in public schools are Hispanic and black. 

* Breakdown of locations give us more insights about race and ethnicity issue. The proportion of white students enrolled in Staten Island takes almost half of the enrollment, while in other Borough especially Bronx, white students only constitute 4.2% of the total enrollment. 

Below we create a random sample of school to create a parallel coordinate plot. You can select, drag and reorder the columns to get a glimpse of race and ethnicity diversity in New York City public school.

```{r parcood}
school <- read.csv("school_f.csv")
sample <- school %>% 
  select(`School.Name`, `X..Asian.1`,`X..Black.1`,`X..White.1`,`X..Hispanic.1`,`X..Multiple.Race.Categories.Not.Represented.1`,location) %>% 
  rename(asian = `X..Asian.1`,
                            black = `X..Black.1`,
                            hispanic = `X..Hispanic.1`,
                            white = `X..White.1`,
                            others = `X..Multiple.Race.Categories.Not.Represented.1`) %>% 
  slice_sample(n = 20) 
parcoords(sample,rownames = F 
    , brushMode = "1D-axes"
    , reorderable = T
    , queue = T
    , color = list(colorBy = "location", colorScale = "scaleOrdinal"
      ,colorScheme = "schemeCategory10"),withD3=TRUE
    )   
```

### Poverty Level in Demographic School

Poverty is  factor worth considering when talking about school diversity.We first tried to capture a top-down view of `Poverty Level` using histogram. Clearly, if the students are random selected to be enrolled in demographic schools, the distribution of their poverty level will be nearly normal distributed.

```{r ch5.7}
pover_dis <- ggplot(type_c, aes(x = `X..Poverty.1`,y =..density..)) + 
  geom_histogram(binwidth = 5,fill = "#9FC29F",colour = "#80593D") +
  geom_density(color = "#80593D",size = 0.8) + 
  xlab("proportion of students in poverty") +
  ylab("")
pover_dis
```

Clearly the distribution of proportion of poverty students are largerly right skewed, which indicates that most of students in demographic schools are those who live in a 